---
title: 'Lab 05: Multiple Comparisons'
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: TRUE
---

In this lab, we will learn how to perform planned contrasts and post hoc analysis for one-way ANOVA.

First, we need an `aov` object. We will continue with the `PlantGrowth` dataset from the previous lab.

```{r}
data("PlantGrowth")
plant.aov <- aov(weight ~ group, data = PlantGrowth)
summary(plant.aov)
```

# Descriptive Statistics

There are multiple functions from many packages that provide descriptive statistics (e.g., group means, *SD*). We will use `psych::describeBY` and `apaTables::apa.1way.table`

```{r}
library(psych)
psych::describeBy(weight ~ group, data = PlantGrowth) # This will give us detailed descriptive stats by group

library(apaTables)
apa.1way.table(dv = weight, iv = group, data = PlantGrowth)
# For apaTables, you can also save a Word document file into the workspace
apa.1way.table(dv = weight, iv = group, data = PlantGrowth, filename = "onewaydesciptive.doc")

# Extra: For APA formatted ANOVA table
apa.aov.table(plant.aov, filename = "anovatable.doc")

```

# Planned contrasts

In *a priori* contrasts (usually just 'contrasts'), we determine a set of comparisons beforehand (i.e., before the data collection). The number of planned comparisons are determined prior to the data collection. Hence, the familiy-wise error rate are known beforehand.

In a modern standard, these specific hypotheses are usually "pre-registered" on a public site (such as the Open Science Framework website: <http://osf.io>).

Recall that in the `PlantGrowth` dataset, we have three conditions: control, treatment 1, treament 2 (in this particular order). The order of levels in a factor is VERY IMPORTANT when analyzing contrasts. **This is the order that you will use in a coefficient matrix.**

```{r}
summary(PlantGrowth$group)

#or
levels(PlantGrowth$group)
```

Suppose that we have two contrasts in our mind. We believe that treatmant 1 and treatment 2 will result in more weight than the control group. In other words, we plan to contrast (`trt1` + `trt2`)/2 with `ctrl`. The other contrast is between `trt1` and `trt2` because we want to know which one is better for plant growth.

In sum, we have *two comparisons* to make (`trt1` + `trt2`)/2 - `ctrl` and `trt1` - `trt2` (or `trt2` - `trt1`, depending on which direction you want to investigate).

## `emmeans` package

To calculate contrasts and post hocs, we will use the `emmeans` (estimated marginal means) package.

```{r}
#install.packages("emmeans")
library(emmeans)
```

First, we need to specify how many comparions do we want and represent each comparison in a coefficient matrix. Each row in the matrix represent level of a factor, in this case, `ctrl`, `trt1`, and `trt2`. It is important to note that an order of coefficients must correspond to levels of a factor. Each column represents our comparisons/contrasts.

```{r}
contrast_m <- data.frame("trt1trt2.vs.ctrl"  = c(-1,  1/2, 1/2), # trt1 and trt2 were averaged (each weight 1/2) to compare against crl (-1)
                         "trt1.vs.trt2"      = c( 0,    1,  -1)) # trt1 (+1) against trt2 (-1). ctrl is leftout (0).
contrast_m
```

Next, we will use `emmeans` to create an emmGrid object, which is an object containing estimated marginal means for each group (i.e., group mean). For this analysis we need two arguments in `emmeans(object, specs)`

-   For `object`, we will use the `aov` object.
-   For `specs`, we will specify that we want means for each group with `~ group`.

```{r}
emmeans(plant.aov, ~ group)
plant.emm <- emmeans(plant.aov, ~ group) # save to an object for later use
```

*Note*: Looking at the means, you might notice that `trt1` was actually lower than `ctrl`. Combining `trt1` and `trt2` will likely cancel each other out. Combining treatment conditions only make sense if they are similar in some aspects. In this case, it is likely that `trt1` and `trt2` are totally different kind of treatment and should not be combined. However, we will proceed with this contrast for a demontration purpose.

## `emmeans::contrast` function

Next we will run contrasts on those group means. The `contrast` function will need four arguments `contrast(object, method, adjust, infer)`

-   `oject` is an emmGrid object from the `emmeans` function.
-   `method` will be our coefficient matrix `contrast_m`.
-   `adjust` is a *p* value adjustment method for multiplicity. Let's use `"bonferroni"`. Some other options are (`"tukey", "scheffe", "sidak", "mvt", "none"`)
-   `infer` is an option for inferential stats. Choose `TRUE` to display both *t* tests and *CI*s.

```{r}
contrast(plant.emm, method = contrast_m, adjust = "none", infer = TRUE) # results with no p value adjustment
contrast(plant.emm, method = contrast_m, adjust = "bonferroni", infer = TRUE) # p values adjusted with bonferroni method. Notice that it multiply each p value by the number of comparisons (2). 
plant.contrast <- contrast(plant.emm, method = contrast_m, adjust = "bonferroni") # save as an object for plotting
```

```{r}
# Plotting contrasts and their confidence interval
plot(plant.contrast)
```

Looking at the results after adjustment, we can see that the `trt1trt2 vs ctrl` contrast was not significant (*p* value above .05 and 95% CI contains zero). That is, when combining `trt1` and `trt2` together, the plant weight was not different from `ctrl`. On the other hand, `trt2` resulted in significantly heavier plants than `trt1`.

## Finding *M* and *SD* for complex contrasts

For a contrast that combine groups together, their mean and *SD* would not be readily available in a regular descriptive statistics table. You will need to extract those groups from the dataset to calclate their means and standard deviations.

```{r}
trt1.trt2 <- subset(PlantGrowth, group == "trt1" | group == "trt2")
psych::describe(trt1.trt2$weight)
ctrl <- subset(PlantGrowth, group == "ctrl")
psych::describe(ctrl$weight)
```

# Post hoc

A *post hoc* analysis is usually done in a pairwise manner (i.e., looking at all possible pairs). Because of a larger number of comparisons, conservative adjustment, such as Bonferroni method, is not recommended. We will use Tukey's Honest Significant Differences (HSD) instead. There are multiple ways to run Tukey's HSD. We will mention the Base R `TukeyHSD` and `emmeans::pairs`.

## Base R

```{r}
TukeyHSD(plant.aov) # input is an aov object. 
plot(TukeyHSD(plant.aov))
```

Among the three groups, only `trt2` was significantly higher than `trt1`. The `ctrl` was to be somewhere in the middle between `trt1` and `trt2` and was not significantly different from either of them.

## `emmeans::pairs`

```{r}
pairs(plant.emm, adjust = "tukey", infer = TRUE) #input is an emm object. Options are similar to contrast, but without `method = `. 
plant.pairs <- pairs(plant.emm, adjust = "tukey") #save for later use. 
```

You can also use `contrast(method = "pairwise")`.

```{r}
contrast(plant.emm, method = "pairwise", adjust = "tukey", infer = TRUE)
```

You can use the `coef` function to look at a coefficient matrix for `"pairwise"` method. You can see how each combinaiton was compared.

```{r}
coef(plant.pairs)
```

And the plot for pairwise comparisons.

```{r}
plot(plant.pairs)
```

# Post hoc for unequal variance

If a homogeneity of variance assumption is violated, you should use Welch's one-way test instead of ANOVA. For post-hoc, you can use Game-Howell Post-hoc test from the `rstatix` package.

```{r}
# install.packages("rstatix")
library(rstatix)
games_howell_test(PlantGrowth, weight ~ group) # input arguments are (data, model)
detach(package:rstatix)
```

# Plotting

To create a plot for a report, `ggplot2` is prefered over the Base R graphic.

## Bar graph

```{r}
plant.summary <- summary(plant.emm) # create a variable containing means and SDs for each condition
plant.summary

plant.summary$Condition <- factor(plant.summary$group, labels = c("Control", "Treatment 1", "Treatment 2")) # create a new factor "Condition" and re-label all levels. 
plant.summary

library(ggplot2)
ggplot(plant.summary, aes(x = Condition, y = emmean)) + #use Condition from plant.summary as X-axis; emmean for Y-axis.
  geom_col(aes(fill = Condition)) + # Add column geometry and fill the color by condition
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL, width = .3)) + # use lower.CL and upper.CL from plant.summary to create error bars. Adjust the width to make them look nice.
  xlab("Condition") + # change X axis label to Condition
  ylab("Weight") + # change Y axis label to Weight
  theme_classic() # classic theme is most similar to APA format. 

ggsave("mean_plot.png") # You can save the graph to a file in a working directory.
```

## Dot plot

```{r}
ggplot(plant.summary, aes(x = Condition, y = emmean, group = 1)) + #similar to above graph, but need `group = 1` option. 
  geom_point() + # Create a point for each mean
  geom_line() + # create a line connecting each group mean
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) + #error bars
  ylim(c(0, 6.5)) + #set Y axis to show 0-6.5 values
  xlab("Condition") +
  ylab("Weight") +
  theme_classic()
```
