---
title: "Lab 13: Independent Samples *t*-Test"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: TRUE
---

# Basic concepts
An independent *t*-test is used to compare means of two groups that are independent from one another (think a between-subject design). For example, one group is an experimental condition, and the other is a control condition. One group is people who endorse a particular political view, the other group is people who disagree with the view. An observation (a sample) can only belong to one group; it cannot belong to both groups at the same time. 

Generally, datasets for independent *t*-tests will have at least two variables: a dependent variable and a grouping (independent) variable. 

Let's consider the dataset `mtcars` in R. We are going to compare whether a car's tranmission (`am`; 0 = automatic, 1 = manual) affects fuel economy `mpg`.

```{r}
data("mtcars") #load dataset from R
df <- mtcars[, c("mpg", "am")] #extract only two columns
df

df$am <- factor(df$am, labels = c("automatic", "manual")) # convert the grouping variable to factor.
df
```

The null hypothesis was that the means of group 1 and group 2 were not different, $H_0: \mu_1 - \mu_2 = 0$

Because the samples in each group were not the same observation, we have to look at the difference at the group level. That is, to look at the difference between $\bar{X}_1$ and $\bar{X}_2$. 

Then, we test whether the mean difference were more or less than zero. $(\bar{X}_1 - \bar{X}_2)-0$

If we take that mean difference and divide it by its standard error, we get a *t* value. 
$$ t = \frac{(\bar{X}_1 - \bar{X}_2)-0}{SE_{\bar{X}_1 - \bar{X}_2}} $$
In a classical Student's *t*-test, to estimate a standard error of a mean difference ($SE_{\bar{X}_1 - \bar{X}_2}$), we need to make an assumption of *homogeneity of variance*. We assume that variances in Group 1 and Group 2 are NOT different. The idea behind this assumption was that if variances of both groups are the same, we can combine (*pooled*) that information together and have a better estimate of standard error. 

A *t*-test with pooled *SD* ($s_p$) looks like this.

$$ t = \frac{\bar{X}_1 - \bar{X}_2}{s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}} $$
The term, $s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$, is an estimated standard error ($SE_{\bar{X}_1 - \bar{X}_2}$).

We can find the pooled *SD* with this equation. 
$$s_p = \sqrt{\frac{(n_1 -1)s_1^2 + (n_2 -1)s_2^2 }{n_1 + n_2 - 2}}$$
Conceptually, we combine the variances from both groups, then take a square root of the variance to get a pooled standard deviation ($s_p$). The degrees of freedom for this test is $(n_1 - 1) + (n_2 -2)  = n_1 + n_2 -2 = N -2$

## Explore data

First, let's look at descriptive stats for each group. We will use `describeBy(y ~ x, data)` from the `psych` package. The function will create descriptive stats of dependent variable `y` by the group variable `x`. 
```{r}
library(psych)
describeBy(mpg ~ am, data = df)
```
We can see that the means of the two groups were not quite the same, but we need a statistical test to determine that. 

# Manual calculation

```{r}
#Mean difference
mean_group <- aggregate(mpg ~ am, data = df, mean) # use aggregate() to apply mean() to `mpg` at each level of `am`
mean_group # this is a dataframe with means values.

mean_diff <- mean_group$mpg[2] - mean_group$mpg[1] #manual - automatic
mean_diff
```
The mean difference is positive, suggesting that Group 2 is higher than Group 1.
```{r}
# Variances
var_group <- aggregate(mpg ~ am, data = df, var) # calculate variances, var(), for each group. 
var_group
var1 <- var_group$mpg[1]
var2 <- var_group$mpg[2]
```
Note that the variances were quite different. We will come back to this issue later. 

```{r}
# Calculate n
n1 <- nrow(df[df$am == "automatic",])
n2 <- nrow(df[df$am == "manual",])
N <- nrow(df)
# pooled standard deviation
s_p <- sqrt((((n1-1)*var1) + ((n2-1)*var2))/(n1 + n2 -2))
s_p

# standard error of the mean difference
se <- s_p*(sqrt(1/n1 + 1/n2))
se

# t-test
t = mean_diff/se
t
```

# `t.test()`

To conduct a Student's *t*-test in R, we use `t.test(y ~ x, var.equal = TRUE)`. The model `y ~ x` denotes a model with `y`, the dependent variable, and `x`, the independent (group) variable. The option `var.equal = TRUE` tells R to use a Student's *t*-test. 

```{r}
t.test(mpg ~ am, data = df, var.equal = TRUE)
```
Note that the *t* value was the same as our manual calculation, but with a negative sign. This is because `t.test()` use *automatic* - *manual*, but our calculation above use *manual* - *automatic*. Nonetheless, they meant the same. 

The 95% CI was for the *mean difference*. Because the 95% CI did not include zero, we were confident to say that the mean differences was not zero, suggesting that the two groups were not the same. 

## Effect size (pooled *SD*)

We use `cohens_d()` from the package `effectsize`.
```{r}
library(effectsize)
effectsize::cohens_d(mpg ~ am, data = df, pooled_sd = TRUE)
```


# Equal variance NOT assumed

When the assumption of homegeneity is uncertain or violated. We should use another version of *t*-test that has been modified for this situation: The Welch *t*-test.
In Welch's formula, the standard error was calculate from each group variance instead of $s_p$. 
$$ t = \frac{\bar{X}_1 - \bar{X}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} $$
The degrees of freedom were then adjusted with the following formula. 
$$ \text{Welch's }df =\frac{\bigg(\cfrac{s_1^2}{n_1} + \cfrac{s_2^2}{n_2}\bigg)^2}   {\cfrac{\bigg(\frac{s_1^2}{n_1} \bigg)^2}{(n_1-1)} + \cfrac{\bigg(\frac{s_2^2}{n_2} \bigg)^2}{(n_2-1)} } $$
Actually the default for `t.test()` is a Welch's *t*-test. 
```{r}
t.test(mpg ~ am, data = df)
```

In this example, the Welch's *t* value was smaller than the Student's *t*. Note that the Welch's *df* was much lower than *N* - 2. This was because of the big difference in *variances* between the two groups. 
```{r}
var_group
```
If we look at the distribution, we will see that the `mpg` in manual transmision had more variation than those in automatic transmission. 

```{r}
library(ggplot2)
ggplot(data = df, aes(x = am, y = mpg)) +
  geom_boxplot() +
  theme_classic()
```
## Effect size (un-pooled *SD*)

```{r}
library(effectsize)
effectsize::cohens_d(mpg ~ am, data = df, pooled_sd = FALSE)
```

# Which option?

Should you use Welch's or Student's *t*? Traditionally, you would, first, determine whether the homogeneity of variance assumption was violated. If it was not, choose Student's *t*. But if it was violated (i.e., unequal variances) like this case, choose Welch's. 

However, when the homogeneity of variance assumption is true, Welch's and Student's *t* produce very similar results. But in the case of violation, Welch's procedure helps protect against Type I error. *Therefore, it is recommended that you should use* **Welch's** *for all cases.* 



# Assumption Check

When conducting independent *t*-test, these assumptions should be checked.

* Independence of observations. Each observation are not related to any other observation. 
* Normality. The values of DV for **each group** should be normally distributed. 
* Homogeneity of variance. The variance of DV are equal in each group. 

The assumption of independence must be ensure by the research design. In this case, we know that the group membership of each observation was mutually exclusive. 

The assumption of normality could be check by Q-Q plots and Shapiro-Wilk statistic. 
For Q-Q plots, the package `car` provides an easy to use function for checking Q-Q plot. 
```{r}
#install.packages("car")
library("car")
qqPlot(mpg ~ am, data = df) #we use a model, mpg ~ am, here to create a QQ plot for each level of `am`.
```
For Shapiro-Wilk test, we will apply the `shaprio.test()` function *by* each group level of `am`. 
```{r}
by(df$mpg, df$am, shapiro.test)
```
The tests for both groups were not significant (*p* > .05), suggesting that `mpg` in each group were normally distributed. 

The assumption of homogeneity of variance could be check with Levene's test. The `car`'s `leveneTest(y ~ x, data, center = mean)` provide a function to test this assumption. 

```{r}
leveneTest(mpg ~ am, data = df, center = mean) #An option `center = mean` uses original Levene's formula.
```
As expected, the test shows that the variances between the two groups were significantly different ($\sigma^2_1 ≠ \sigma^2_2$), which violated the homogeneity of variance assumption. The Welch's procedure is preferred in this case. 

## `jmv`
The jamovi's R `jmv` package provides another way to do *t*-tests in R with `ttestIS()`
```{r}
library(jmv)
ttestIS(formula = mpg ~ am, data = df,
        students = TRUE, #Student's t-test
        welchs = TRUE, #Welch's t-test
        eqv = TRUE, #Levene's test
        meanDiff = TRUE, #mean difference and SE
        desc = TRUE #descriptive stats
        )
```

