---
title: 'Lab 06: Visualizing Distribution'
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set working directory to a source file.  
``setwd(dirname(rstudioapi::getSourceEditorContext()$path))`  
Then, load the data.

```{r import data}
solar_data <- read.csv("data/RES_STAT_Lab6_Data.csv")
solar_data$inno <- factor(solar_data$inno, levels = c("innovators", "early adopters", "early majority", "late majority", "laggards"), ordered = TRUE)
head(solar_data$inno)
```

# Boxplot
A boxplot is a classic visualization of a distribution. You can run the function in the base R or with the `ggplot2` package.
## Base R

```{r boxplot}
boxplot(solar_data$attitude)
```

You can also assign an object name for your plot.
```{r boxplot object}
bp <- boxplot(solar_data$attitude) #This still prints out the plot.
bp
```
The object `bp` is a list containing information about the plot.  
`$stat` contains 1) lower whisker or -1.5 IQR, 2) lower hinge or 25th quantile, 3) median or 50th quantile, 4) upper hinge or 75th quantile, and 5) upper whisker or +1.5 IQR.
`$n` is the number of observations.
`$out` list are values of outliers that lie outside the whiskers
See `help(boxplot)` for more information.

The values in boxplot object can be access as a list. This can be useful. For example, we could use `$out` to identify outliers data in the dataset.
```{r find outliers}
solar_data[solar_data$attitude %in% bp$out, ] 
#%in% is a value matching function. We match value in $attitude with the outliers, $out.
```

### Multiple Boxes
When exploring the data, you can use boxplot to simultaneously check  distributions of multiple items at once. 
```{r}
boxplot(solar_data[, c("att1", "att2", "att3", "att4", "att5")]) #check distributions of item att1-5
```

### Boxplots by group
You can use formula to create boxplots by each level of a factor. In the formula `attitude ~ inno`, `attitude` is split by the factor `inno`. The `solar_data` is the dataset of variables in a formula.  
`xlab` and `ylab` are labels for x- and y-axis, respectively.

```{r}
boxplot(attitude ~ inno, solar_data, xlab = "Innovator Groups", ylab = "Attitude Score")
```


## Boxplot with `ggplot2`
While the base R is simple, `ggplot2` offers a more coherent way of writing a code for graphics. It employs the "*Grammar of Graphics*." The main components are 1) a **data**, 2) **aesthethics**, and 3) **geometries**. 

Data is the data you want to plot.
Aesthetics `aes()` is how you want to map the data to the plot scale. 
Geommetries (geom) is the visual elements of the plot (e.g., lines, dots, bars)

There are other option layers, such as *themes*, *statistics*, *coordinates*, and *facet*. The graphics are built by the multiple layers.

The function `ggplot` is the main function to plot graphics. The first argument is the data `solar_data`. The second argument is aesthetics `aes()`. Inside we specify `y = attitude`, meaning that we want to plot the value on *attitude* on the y-axis. After that, we use a `+` sign to tell the function that we still want to add more layers.   
Then we add a geometry layer `geom_boxplot()` to create a boxplot. Then `+` the theme layer `theme_classic()`.

```{r}
library(ggplot2) #load the ggplot2 package
ggplot(solar_data, aes(y = attitude)) +
  geom_boxplot() +
  theme_classic() #Classic theme is the most similar to the APA style. 
```

### Boxplot by groups
You can specify a factor `x` in `aes()` to create boxplots by groups. In this case, we use `inno` for grouping on the *x*-axis. 

```{r box plot by group}
ggplot(solar_data, aes(x = inno, y = attitude)) +
  geom_boxplot() +
  theme_classic()
```


# Histogram
If you want to see the shape of a distribution, you will need a histogram. 

## Base R
```{r histrogram}
h <- hist(solar_data$attitude)
h
```
The object `h` is a list of values for a histogram. 
`breaks` = the boundary of bins.
`counts` = frequencies for each bin.
`density` = estimated density values.
`mid` = midpoints for each bin.
`equidist` = logical, if the bin sizes are equal. 

R will automatically decide how to bins look like. If you want your histogram to look a certain way, you will use the argument `breaks`.
There are multiple ways to use the argument `breaks` (see Help). We will use a vector to determine the boundary of the bins. 

```{r customize histrogram}
# First create a vector from 0 to 4 with a 0.25 step.
b <- seq(0, 4, 0.25)
b
# Then use b in the argument breaks.
hist(solar_data$attitude, breaks = b) # alternative: hist(solar_data$attitude, breaks = seq(0, 4, 0.25))


# use col for color and ylim for the range of y-axis.
hist(solar_data$attitude, breaks = seq(0, 4, 0.25), col = "lightsteelblue", ylim = c(0, 80))
```

See http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf for color names in R. 

## Histogram with `ggplot2`
For histogram `aes()`, we use `x = attitude` as we plot the value of attitude on the *x*-axis. Then we add a layer of `geom_histogram`.  
By default, it will use 30 bins. 
```{r histogram}
ggplot(solar_data, aes(x = attitude)) +
  geom_histogram() +  
  theme_classic() 
```

We can specify a bin width using `binwidth`
```{r}
ggplot(solar_data, aes(x = attitude)) +
  geom_histogram(binwidth = 0.5) +
  theme_classic() 
```

We can also change a border and a filll color with `color` and `fill`.
```{r}
ggplot(solar_data, aes(x = attitude)) +
  geom_histogram(binwidth = 0.5, color = "red", fill = "pink") + 
  theme_classic()
```

### Histogram by groups

Because historgram is plot on the x-axis, if we want to split histogram by groups, we need to create multiple of them and put them to gether using a facet layer `facet_wrap()`. 

```{r facet wrap}
ggplot(solar_data, aes(x = attitude)) +
  geom_histogram(binwidth = .5,
                 aes(y = ..density..), #plot density instead of frequency
                 color = "black",
                 fill = "steelblue") + 
  facet_wrap(~ inno) + # create subplots by inno and put them together
  theme_classic()
```

# Density Plot
Density plot shows the probability density of the distribution.
## Base R
```{r density plot}
plot(density(solar_data$attitude))
```

## `ggplot2`

```{r ggplot density}
ggplot(solar_data, aes(attitude)) +
  geom_density(color = "red", fill = "pink") +
  theme_classic()
```

We can overlay the density plot on the histogram plot by layer them together.
```{r ggplot histogram with density}
ggplot(solar_data, aes(attitude)) +
  # histogram layer
  geom_histogram(binwidth = .25, 
                 aes(y = ..density..), #histogram of density
                 color = "black", 
                 fill = "white") +
  # density plot layer
  geom_density(alpha = .2, color = "forestgreen", fill = "forestgreen") + #alpha = transparent level
  theme_classic() +
  facet_wrap(~ inno)
```

# Violin Plot
A violin plot is a hybrid between a density and a boxplot. 
## Violin with `ggplot2`
Violin plots are usualy used to compare distributions for each leve of a factor. For the `aes()`, we use `x = inno` to split plots by innovator groups. `y = attitude` to plot the distribution on the *y*-axis. `fill = inno` to fill color automatically by innovator groups. 
```{r violin plot}
ggplot(solar_data, aes(x = inno, y = attitude, fill = inno)) +
  geom_violin() +
  theme_classic()
```

## Violin plot with data points
We can add `geom_jitter` layer to illustrate how data points distribute in the plot. 
```{r violin plot with data}
ggplot(solar_data, aes(x = inno, y = attitude, fill = inno)) +
  geom_violin() +
  geom_jitter(width = .15) + #width = 15% to keep most dots in each violin.
  theme_classic()
```

# Resource: ggplot in datacamp.


