---
title: "Lab 05: Standard Scores"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: TRUE
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Data

Save a data file and an R script in the same directory.
Set working directory to Source File Location.
`setwd(dirname(rstudioapi::getSourceEditorContext()$path))` Then, load the data.

```{r import data}
solar_data <- read.csv("data/RES_STAT_Lab5_Data.csv")
head(solar_data) ## view the first part of the data frame.
```

# Data Preparation

ในแบบฝึกหัดนี้เราจะใช้ตัวแปรความรู้เกี่ยวกับแผงโซลาเซลล์ (knowledge) ตัวแปรนี้ได้มาจากการเฉลี่ยคะแนนการประเมินความรู้ของตนเองจำนวน

# Calculate *z*-scores

For each *X*, its standardized score, *z*, can be calculated as $$z = \frac{X-\bar{X}}{s}$$

### Manual method

You can calculate *z*-score step by step.
First, the top part of the equation $X-\bar{X}$ is called **mean centering**.
That is, the value X is subtracted by the mean of X, `X - mean(X)`.\
The mean of X is now the "center" (0) of the distribution.
Each $X-\bar{X}$ represents how far each X is from the $\bar{X}$, i.e., a deviation score.

Then, the centered value is **scaled** to the SD of X, `X - mean(X)/sd(X)`.
When the score is scaled to SD.
It means that one unit change in the *z*-score represents one SD change in the raw score.

To standardize a variable, you will calculate *z*-scores for each row in that column.
Let put it the calculated values in `z_kno_m` variable.\
`z` for *z*-scores; `kno` for knowledge; and `m` for manual calculation.

```{r manual z score}
solar_data$z_kno_m <- (solar_data$knowledge - mean(solar_data$knowledge))/sd(solar_data$knowledge)
head(solar_data$z_kno_m)
```

### `scale()` function

The scale function will *center* the value by substracting the column mean.
Then it *scales* the centered value by the column standard deviation.\
Let create a variable name `z_kno_f`. `f` for function.

```{r scale function}
solar_data$z_kno_f <- scale(solar_data$knowledge)
head(solar_data$z_kno_f)
```

Are all values in the manual method equal to values from `scale()`?

```{r compare manual and scale}
all(solar_data$z_kno_m == solar_data$z_kno_f)
```

Both methods lead to the same results.

# Outliers

### Use *z*-score to determine outliers

There are many methods to identify outliers.
In this example, any values outside ±3 SD are considered outliers.
Because *z*-scores are in a unit of SD, any *z* beyond ±3 will be marked as outliers.

```{r all outliers index}
outliers <- solar_data$z_kno_f > 3 | solar_data$z_kno_f < -3
solar_data[outliers, ]
```

### *z*-score by group

`z_kno_f` was calculated from the entire data set.
This would be appropriate if we are looking at a homogenous group.
However, it is not a case here.
Our data contain subgroups of innovators `inno`.
Therefore, we should find outilers *within* each group instead.

We will use `ave()` to calculate the z-score by group.
`ave` is normally used to calculate means for subsets of X.
However, you can specified the argument `FUN =` to other functions, such as `sd()` or `scale()`.

The first argument in `ave()` is a numeric object to be calculated (`knowledge`).
The second argument is a grouping variable (`inno`).
The third argument `FUN =` is a function to by applied for each factor level (`FUN = scale`).

```{r z scores by group}
solar_data$z_kno_g <- ave(solar_data$knowledge, solar_data$inno, FUN = scale)
# wrong <- ave(solar_data$z_kno_f, solar_data$inno, FUN = scale)
head(solar_data$z_kno_g)
# head(wrong)
```

To check whether the code works correctly, we will calculate the mean and SD of the *z*-scores for each innovator group.
The mean should be 0 and SD should be 1 for each group.

We will use `tapply` to apply `mean()` to each `inno` group.
In `tapply`, the first argument is an input value.
The second argument is a grouping variable (a factor).
The third argument is a function to apply.

```{r mean and sd by group}
tapply(solar_data$z_kno_g, solar_data$inno, mean)
# The values contain very small decimals. Let's round them up to make them easier to read.
round(tapply(solar_data$z_kno_g, solar_data$inno, mean))
# Then calculate the SD
tapply(solar_data$z_kno_g, solar_data$inno, sd)
```

With the mean = 0 and SD = 1 for each innovator group, it seems that the *z-score-by-group* code works correctly.

### Outliers by group

Now we find outliers that have z-score beyond ±3.
Because the *z*-scores were calculated for each innovator group, any values beyond ±3 are outliers for that group.

```{r outliers by group}
outliers_g <- solar_data$z_kno_g > 3 | solar_data$z_kno_g < -3
solar_data[outliers_g, ]
```

In this case, we will remove outliers from the data.
We will choose rows that are **NOT** outliers `!outliers_g`.

```{r}

solar_data_new <- solar_data[!outliers_g, ] # choose rows that are not outliers and choose all columns.
nrow(solar_data) # Number of observation in the original data. 
nrow(solar_data_new) # The number of observation should be reduced by 2 cases.
```

# Descriptives for each group

We will use `describe()` from package `psych` to calculate means and SDs `by` each `inno` group.
The first argument of `by` is an input data, `knowledge`.
The second argument is a grouping factor, `inno`.
The third argument is a function to apply, `describe`.

```{r}
library(psych)
by(solar_data_new$knowledge, solar_data_new$inno, psych::describe)
# We will create an object from the code above for future use.
kno_table <- by(solar_data_new$knowledge, solar_data_new$inno, psych::describe)
```

The object `kno_table` from `by()` and `describe()` is a list.
We will access its mean and SD values to calculate effect sizes.

# Effect sizes

An effect size is $\frac{X_1-X_2}{SD}$.

We will need means of the two groups and an SD.

```{r}
kno_table$innovators$mean # mean of the innovators group
kno_table$laggards$mean # mean of the laggards group
# For SD we will use the SD from the whole data set.
sd_all <- sd(solar_data_new$knowledge)
sd_all

```

Now we can calculate an effect size for each comparison.

```{r}
effect_size1 <- (kno_table$innovators$mean - kno_table$laggards$mean)/sd_all
effect_size1
effect_size2 <- (kno_table$innovators$mean - kno_table$`early adopters`$mean)/sd_all
effect_size2
effect_size3 <- (kno_table$`early majority`$mean - kno_table$laggards$mean)/sd_all
effect_size3
effect_size4 <- (kno_table$`early majority`$mean - kno_table$`late majority`$mean)/sd_all
effect_size4

```
